---
title: "Working with categorical predictors"
author: "Matthew Rudd"
date: "3/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The primary message in the first example is:

> ***Categorical data is not quantitative!***

# Logistic regression: violet data  

Load the violet data and compute the bud formation rates for each plot:

```{r message=FALSE} 
library(tidyverse)
violets <- read_csv("violets.csv")
violets <- violets %>%
  mutate(Bud = as.numeric( `Bud counts` > 0 ))

(violet_plots <- violets %>%
  group_by(Plot) %>% 
  summarize( N = n(),
             Rate = mean(Bud)) )
```

The different plots are numbered 1 through 10; by default, R therefore 
considers `Plot` to be a numeric variable, which we can confirm by checking its
type:

```{r}
class(violets$Plot)
```

What could go wrong?

## Useless model when `Plot` is numeric

If we fit a logistic model now, using `Plot` as a _quantitative_ variable, the results
are terrible:

```{r}
violet.fit <- glm( Bud ~ Plot, data=violets, family="binomial")
summary( violet.fit )

violet_plots <- violet_plots %>%
  mutate( Fit = predict( violet.fit, 
                         newdata = data.frame(Plot = 1:10),
                         type = "response"))

violet_plots
```

The fitted values, which should correspond to the probabilities of bud formation
in each plot, are not good estimates of the empirical rates.

## Reasonable model when `Plot` is categorical

Once we recognize that `Plot` is categorical and convert it into a factor, 
the resulting logistic model works well:

```{r}
violets$Plot <- as.factor( violets$Plot )
violet.fit <- glm( Bud ~ Plot, data=violets, family="binomial")
summary( violet.fit )

violet_plots <- violet_plots %>%
  mutate( Fit = predict( violet.fit, 
                         newdata = data.frame(Plot = levels(violets$Plot)),
                         type = "response"))

violet_plots
```

This is what happens when working with one categorical predictor:

> ***the model predictions are the group averages.***

In this case, the average for a group (a plot) is the empirical probability
of bud formation, i.e., the percentage of days on which new buds were observed in 
a given plot.

# Linear regression: infant mortality rates by continent

## Visualize the data and examine the summary stats

```{r}
imr <- read_csv("imr2011.csv")
imr$Continent <- as.factor(imr$Continent)
imr <- imr %>%
  mutate( Continent = fct_reorder(Continent, -IMR))

ggplot( imr, aes( x=IMR, y=Continent, color=Continent)) +
  geom_boxplot() +
  geom_point() +
  theme( legend.position = "none")

imr_means <- imr %>% 
  group_by(Continent) %>%
  summarize( Mean = mean(IMR)) 

imr_means %>%
  arrange(Mean)
```

### Linear regression model

```{r}
imr.fit <- lm( IMR ~ Continent, data = imr )
summary(imr.fit)

imr_means <- imr_means %>%
  mutate( Fit = predict( imr.fit,
                         newdata = data.frame(Continent = levels(imr$Continent))))
imr_means %>% arrange(Mean)
```

# Linear regression with 2 predictors

```{r}
heights <- read_csv("all_heights.csv")

ggplot( heights, aes(x=Parent, y=Child, color=Sex)) +
  geom_point()

ggplot( heights, aes(x=Parent, y=Child, color=Sex)) +
  geom_point() +
  geom_smooth( method="lm", level=0)

# no interaction
heights.fit <- lm( Child ~ Parent + Sex, data=heights)
summary(heights.fit)

#with interaction
heights.fit <- lm( Child ~ Parent*Sex, data=heights)
summary(heights.fit)

```