---
title: "Chronic conditions and logistic regression"
author: "Matthew Rudd"
date: "2/10/2022"
output: html_document
---

# The 2009 NAMCS and chronic conditions

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(pander)
library(flextable)
chronic <- read_csv("chronic.csv")
```

## The null model

```{r}
chronic.null <- mean( chronic$Condition) 

chronic %>% 
  group_by(Condition)  %>% 
  summarize( N=n(), Percentage = N/nrow(chronic))
```

## A useless plot 

```{r}
ggplot( chronic, aes(Age, Condition)) + geom_point()
```

## Some useful plots

```{r}
cohorts <- chronic %>% 
  group_by(Age) %>% 
  summarize( N=n(), Percentage = sum(Condition)/N )

ggplot( cohorts, aes(Age, Percentage)) + geom_point() + ylim(0,1)

chronic <- chronic %>%
  mutate( Decade = 10*round(Age/10))

decades <- chronic %>%
  group_by( Decade ) %>%
  summarize( N=n(), Percentage = sum(Condition)/N )

# using knitr's kable function :
# knitr::kable( decades )

# using pander :
# pander( decades )

# using flextable's regulartable function :
regulartable( decades ) %>% autofit()

ggplot( cohorts, aes(Age, Percentage)) + 
  geom_point() + 
  geom_point( data=decades, aes(Decade, Percentage), color="tomato", size=3) + 
  ylim(0,1)

```

## The curve of chance

```{r} 
z <- seq(-5, 5, len=1000)
logistic <- data.frame( z=z, sigma = 1 / (1 + exp(-z)))
ggplot( logistic, aes(z,sigma)) + 
  geom_line( size=1 ) +
  labs( title = "The logistic curve")
```

This is a plot of the ***logistic curve*** a.k.a. the ***logistic function*** a.k.a. the 
***sigmoid function*** 

$$
\sigma( z ) ~ = ~ \frac{ 1 }{ 1 + e^{-z}} \ .
$$ 

# Fitting the logistic curve to our data

```{r}
chronic.fit <- glm( Condition ~ Age, data=chronic, family="binomial")
chronic.fit 
```

The ***fitted logistic regression model*** is therefore 

$$ 
\text{Probability of condition} ~ = ~ \sigma( -2.03938 + 0.05182 \times \text{Age}) \ .
$$ 

To phrase this in terms of the odds of having a condition, recall the definition
of the odds of something happening:

$$
\text{Odds of event} ~ = ~ \frac{ \text{Probability of event} }{  1 - \text{Probability of event} } \ .
$$
In terms of ***odds***, our logistic model is 

$$
\log{ ( \text{Odds of a condition} ) } ~ = ~ -2.03938 + 0.05182 \times \text{Age} \ .
$$

```{r}
ggplot( chronic, aes(Age, Condition)) + 
  geom_smooth( method="glm", method.args=list(family="binomial")) +
  geom_point( data=cohorts, aes(Age, Percentage))

new <- data.frame( Age = seq(-70,150,by=1))
extended <- predict( chronic.fit, new, type="response")
extended <-data.frame( Age = new$Age, Probability = extended )

ggplot( chronic, aes(Age,Condition)) +
  geom_point( data=cohorts, aes(Age, Percentage)) +
  geom_line( data = extended, aes(Age, Probability)) +
  xlim(-70,150)

```

## The odds ratio

One way to summarize this model is to report its ***odds ratio***, which is the logistic
regression analogue of the slope in linear regression. To derive the expression for the 
odds ratio, we first need to solve the equation

$$
y ~ = ~ \sigma(z) ~ = ~ \frac{ 1 }{ 1 + e^{-z}} 
$$
for $z$ ; doing the necessary manipulation yields

$$
z ~ = ~ \log{ \left( \frac{ y }{ 1 - y } \right) } \ .
$$

This inverse function of the logistic function is known as the ***logit*** : by definition, 

$$
\operatorname{logit}( y ) ~ = ~ \log{ \left( \frac{ y }{ 1 - y } \right) } \ .
$$

Putting this all together, a simple logistic model with predictor variable $x$ has two equivalent forms:

$$ 
\text{Probability of event} = \sigma( a + bx ) \quad \iff \quad \log{(\text{Odds of event})} = a + bx \ .
$$

Exponentiating the second form yields yet another formulation:

$$
\text{Odds of event} ~ = ~ e^{a+bx} ~ = ~ e^{a} e^{bx} \ .
$$
Now we get to the odds ratio. Fix a value $x_{0}$ of the predictor and compare the odds when $x=x_{0}$
to the odds when $x=x_{0}+1$:

$$
\text{Odds when} \left\{ x=x_{0}+1 \right\} ~ = ~ e^{a} e^{bx_{0}} e^{b} ~ = ~ e^{b} \times \text{Odds when} \left\{ x=x_{0} \right\} \ .
$$

The odds get multiplied by the factor $e^{b}$, which is, by definition, the model's ***odds ratio***.

> The odds ratio tells us how the odds change when the predictor increases by 1 unit: *multiply* the odds by the odds ratio.

For the logistic chronic condition model:

```{r}
chronic.or <- exp( chronic.fit$coefficients[2] )
chronic.or
```

The odds ratio is 1.053181, so the odds of having a chronic condition increase by about 5% 
with each 1 year increase in `Age`. A 10-year increase in `Age` therefore leads to 
a multiplier of 

```{r}
chronic.or^10
```

i.e., a 68% increase in the odds of having a chronic condition. 
A 50-year increase in `Age`  leads to a multiplier of 

```{r}
chronic.or^50
```

so the odds of a random 70-year-old having a chronic condition are more than 13 times the odds 
of a random 20-year-old having a chronic condition.  Ah, the golden years after retirement!

# Binary classification

## Confusion matrices for varying thresholds

```{r}

# decision boundary where probability = 50% (i.e., odds = 1) :
(threshold <- ( -chronic.fit$coefficients[1] ) / ( chronic.fit$coefficients[2] ))

chronic <- chronic %>%
  mutate( Classifier = as.numeric( Age > threshold ))

# confusion matrix for assessing classifier:
class.table <- table( chronic$Condition, chronic$Classifier)
rownames( class.table ) <- c("No condition", "Has a condition")
colnames( class.table ) <- c("No condition predicted", "Condition predicted")
class.table
# overall accuracy rate
mean( chronic$Condition == chronic$Classifier)
# overall error rate 
mean( chronic$Condition != chronic$Classifier)

chronic <- chronic %>%
  mutate( Classifier = as.numeric( Age >= 80 ))

# confusion matrix for assessing classifier:
class.table <- table( chronic$Condition, chronic$Classifier)
rownames( class.table ) <- c("No condition", "Has a condition")
colnames( class.table ) <- c("No condition predicted", "Condition predicted")
class.table
# overall accuracy rate
mean( chronic$Condition == chronic$Classifier)
# overall error rate 
mean( chronic$Condition != chronic$Classifier)

chronic <- chronic %>%
  mutate( Classifier = as.numeric( Age >= 20 ))

# confusion matrix for assessing classifier:
class.table <- table( chronic$Condition, chronic$Classifier)
rownames( class.table ) <- c("No condition", "Has a condition")
colnames( class.table ) <- c("No condition predicted", "Condition predicted")
class.table
# overall accuracy rate
mean( chronic$Condition == chronic$Classifier)
# overall error rate 
mean( chronic$Condition != chronic$Classifier)

```

### Receiver Operating Characteristic (ROC) curve

```{r}

# pROC and ROCR are the 2 standard R packages for computing/plotting ROC curves

library(pROC)
chronic <- chronic %>% mutate(Probability = chronic.fit$fitted.values )
roc.obj <- roc(chronic$Condition,  chronic$Probability, plot=TRUE, print.auc=TRUE)

```
