---
title: "Count regression"
author: "Matthew Rudd"
date: "Stat 214, Easter 2022"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
# Basic knitr options
library(knitr)
opts_chunk$set(
  comment = NA,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = TRUE,
  cache = FALSE
)
library(kableExtra)
library(tidyverse)
library(gridExtra)
library(DT)
library(MASS)
options(scipen = '999')
```

# How can we relate parameters to predictors?

```{r}
crashes <- read_csv("crashes.csv")
ggplot( crashes, aes(x=pop2010, y=crashes)) +
  geom_point()
crashes$pop2010 <- crashes$pop2010 / 1000
crashes <- crashes %>%
  mutate(pop_group = 9*round(pop2010/9))
ggplot( crashes, aes(x=pop2010, y=crashes)) +
  geom_point()
ggplot( crashes %>% filter(pop2010<90), aes(x=pop2010, y=crashes)) +
  geom_point()
crashes <- crashes %>% filter( pop2010 < 90 )
(crash_means <- crashes %>% 
  group_by( pop_group ) %>%
  summarize( Mean = mean(crashes)))
ggplot( crashes, aes(x=pop2010, y=crashes)) +
  geom_point() + 
  geom_point( data=crash_means, aes(x=pop_group, y=Mean),color="darkorchid",size=4) +
  geom_line( data=crash_means, aes(x=pop_group, y=Mean),color="darkorchid", size=2)
```

Based on this, it seems reasonable to fit a model of the form
$$ \log( \mu ) ~ = ~ a + bx \ , $$
where $\mu$ is the mean response and $x$ is the one and only (quantitative) predictor.
An important consideration, though, is whether the response seems to follow the
Poisson distribution or the negative binomial distribution -- i.e., are its conditional 
mean and variance roughly the same or not?

# Poisson regression

```{r}
births <- read_csv("births.csv")
births$Pop2010 <- births$Pop2010 / 1000
births_by_county <- births %>%
  group_by(county) %>% 
  summarize( Mean = mean(births),
             Variance = var(births),
             Ratio = Variance/Mean ) %>% 
  arrange( desc(Ratio))

datatable( births_by_county, caption = "Monthly birth statistics for U.S. counties")

poisson_births <- births_by_county %>% 
  filter( abs( Ratio - 1 ) < 0.1 ) %>%
  pull( county )
poisson_births <- births %>%
  filter( county %in% poisson_births )

ggplot( poisson_births, aes( x=Pop2010, y=births)) +
  geom_point()

# the best fitting smooth curve is not exponential!
ggplot( poisson_births, aes( x=Pop2010, y=births)) +
  geom_point() +
  geom_smooth() 

# but the fitted Poisson regression curve is!!
ggplot( poisson_births, aes( x=Pop2010, y=births)) +
  geom_point() +
  geom_smooth( method="glm", method.args=list(family="poisson")) 

poisson_births <- poisson_births %>%
  group_by(county) %>%
  mutate( Mean = mean(births)) 

ggplot( poisson_births, aes( x=Pop2010, y=births)) +
  geom_point() +
  geom_smooth( method="glm", method.args=list(family="poisson")) +
  geom_point( aes(x=Pop2010, y=Mean), color="red", size=3)
```

```{r}
# don't use repeated measurements!!!!
dec19 <- poisson_births %>% 
  filter( month=="December" & year==2019)
dec19_null <- glm( births ~ 1, data=dec19, family="poisson")
summary( dec19_null )
dec19_fit <- glm( births ~ Pop2010, data=dec19, family="poisson")
summary(dec19_fit)

ggplot( dec19, aes( x=Pop2010, y=births)) +
  geom_point() +
  geom_smooth( method="glm", method.args=list(family="poisson"))
  # geom_point( aes(x=Pop2010, y=Mean), color="red", size=3)

new_data <- data.frame( Pop2010 = c(125,250))
(predictions <-predict( dec19_fit, newdata=new_data, type="response" ))

# how to sample from a Poisson distribution:
simulation <- rpois( n=20, lambda=predictions[1])
hist( simulation )

# compare observed and simulated births...
# in Paulding County, GA (pretty good) :
new_data <- data.frame( Pop2010 = c(142.324, 268.1))
(lambda <-predict( dec19_fit, newdata=new_data, type="response" ))

paulding <- poisson_births %>%
  filter( county == "Paulding County, GA")
paulding <- paulding %>%
  mutate( simulation = rpois( n=nrow(paulding), lambda=lambda[1]))
ggplot( paulding %>% dplyr::select( births, simulation) %>% pivot_longer(births:simulation), 
        aes(value)) +
  geom_histogram() +
  facet_wrap(~name, ncol=1) +
  labs( title="Paulding County, GA")

# and in Dauphin County, PA (not so good)
dauphin <- poisson_births %>%
  filter( county == "Dauphin County, PA")
dauphin <- dauphin %>%
  mutate( simulation = rpois( n=nrow(dauphin), lambda=lambda[2]))
ggplot( dauphin %>% dplyr::select( births, simulation) %>% pivot_longer(births:simulation), 
        aes(value)) +
  geom_histogram() +
  facet_wrap(~name, ncol=1) +
  labs( title="Dauphin County, PA")
```

# Negative binomial regression 

## A bad model...

```{r}
dec19_all <- births %>% 
  filter( month=="December" & year==2019)
ggplot( dec19_all, aes( x=Pop2010, y=births)) +
  geom_point() +
  geom_smooth( method="glm.nb" )

dec19_all_fit <- glm.nb( births ~ Pop2010, data=dec19_all)
summary( dec19_all_fit )

jefferson <- births %>% 
  filter( county == "Jefferson County, AL")
new_data <- data.frame( Pop2010 = 658.466)
mu <- predict( dec19_all_fit, newdata = new_data, type="response")
jefferson <- jefferson %>% 
  mutate( simulation = rnbinom( nrow(jefferson), mu=mu[1], size=dec19_all_fit$theta))

ggplot( jefferson %>% dplyr::select( births, simulation) %>% pivot_longer(births:simulation), 
        aes(value)) +
  geom_histogram() +
  facet_wrap(~name, ncol=1) +
  labs( title="Jefferson County, AL")
```

## A possibly better model...

```{r}
dec19_all$Pop2010 <- log( dec19_all$Pop2010 )
ggplot( dec19_all, aes( x=Pop2010, y=births)) +
  geom_point() +
  geom_smooth( method="glm.nb" )

dec19_all_fit <- glm.nb( births ~ Pop2010, data=dec19_all)
summary( dec19_all_fit )

jefferson <- births %>% 
  filter( county == "Jefferson County, AL")
new_data <- data.frame( Pop2010 = log(658.466))
mu <- predict( dec19_all_fit, newdata = new_data, type="response")
jefferson <- jefferson %>% 
  mutate( simulation = rnbinom( nrow(jefferson), mu=mu[1], size=dec19_all_fit$theta))

ggplot( jefferson %>% dplyr::select( births, simulation) %>% pivot_longer(births:simulation), 
        aes(value)) +
  geom_histogram() +
  facet_wrap(~name, ncol=1) +
  labs( title="Jefferson County, AL")

```

## Using a factor is not as good...

```{r}
dec19_all <- births %>% 
  filter( month=="December" & year==2019)
dec19_all <- dec19_all %>%
  mutate( PopGroup = cut(Pop2010, breaks=12, labels=1:12 ))
ggplot( dec19_all, aes( x=PopGroup, y=births)) +
  geom_point() +
  geom_smooth( method="glm.nb" )

dec19_all_fit <- glm.nb( births ~ PopGroup, data=dec19_all)
summary( dec19_all_fit )

jefferson <- births %>% 
  filter( county == "Jefferson County, AL")
new_data <- data.frame( PopGroup = "1" )
mu <- predict( dec19_all_fit, newdata = new_data, type="response")
jefferson <- jefferson %>% 
  mutate( simulation = rnbinom( nrow(jefferson), mu=mu[1], size=dec19_all_fit$theta))

ggplot( jefferson %>% dplyr::select( births, simulation) %>% pivot_longer(births:simulation), 
        aes(value)) +
  geom_histogram() +
  facet_wrap(~name, ncol=1) +
  labs( title="Jefferson County, AL")

```
