---
title: "violet"
author: "Hoang Le"
date: "2/22/2022"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(tidyverse)
violet <- read_csv("https://raw.githubusercontent.com/mbrudd/site/master/stat214/data_code/violets.csv") %>% rename(Type = 'Bud type', Count = 'Bud counts')
View(violet)
```

### Bud count - binary response variable for Chasmogamus

```{r}
chasmogamous <- violet %>%
  filter(Type == 'Chasmogamous') %>%
  select(Photoperiod, Count) %>%
  mutate(Bud = as.integer(Count > 0) )

View(chasmogamous)
```

```{r}
cohorts <- chasmogamous %>%
  group_by(Photoperiod) %>%
  summarise(N = n(), 
            Percentage = sum(Bud)/N)

 View(cohorts)

ggplot(chasmogamous, aes(Photoperiod, Bud)) + 
  geom_smooth(method="glm", method.args = list(family="binomial")) +
  geom_point(data=cohorts, aes(Photoperiod, Percentage))

ggplot(chasmogamous, aes(Photoperiod, Bud)) + 
  geom_smooth(method="glm", method.args = list(family="binomial")) +
  geom_point()
```
```{r}
bud.fit <- glm( Bud ~ Photoperiod, data=chasmogamous, family="binomial")
bud.fit
```
```{r}
# decision boundary where probability = 50%, (i.e. odd = 1):
threshold <- -bud.fit$coefficients[1] / bud.fit$coefficients[2]
threshold


chasmogamous <- chasmogamous %>%
  mutate( Classifier = as.numeric(Photoperiod < threshold))
# confusion matrix 
class.table <- table( chasmogamous$Bud, chasmogamous$Classifier)
rownames(class.table) <- c("No bud", "Bud")
colnames(class.table) <- c("No bud predicted", "Bud predicted")
class.table
```

$$
(\text{Wald statistic}) \\
z = \frac{\text{observed} - \text{expected}}{\text{SE}} 
$$
### Likelihood ratio test, to test how much better is this model compared the null model

$$
\
\text{null deviance} - \text{residual deviance} \\
= -2\log(\text{null likelihood}) - (-2\log(\text{model likelihood})) \\
= -2\log(\frac{\text{null likelihood}}{\text{model likelihood}}) = G^2
$$
```{r}
g <- bud.fit$null.deviance - bud.fit$deviance
g
pchisq(g, 1, lower.tail = FALSE) # p value 
```
### Bud count - binary response variable for Chasmogamus
### how bud formation relate to photo-period, use glm, plot stuff --> should get a nice fitting 

### Linear regression
$$
R^2 = \frac{\sum(y_i - \bar{y})^2 - \sum(y_i - \hat{y})^2}{\sum(y_i - \bar{y})} = \frac{\sum(\hat{y_i} - \bar{y_i})^2}{\sum(y_i - \bar{y})^2}= 1 - \frac{\text{RSS}}{\text{TSS}}
$$
### McFadden's pseudo-R^2 --> generalize from the linear regression model
$$
R_M^2 = 1 - \frac{\text{deviance}}{\text{null deviance}}
$$
$$
R_M^2 = 1 \rightarrow \text{logistic model no better then null model}
$$
$$
0.2 < R_M^2 < 0.4 \rightarrow \text{good model}\\
R_S^2 = \frac{\sum_i(\hat{\pi_i} - \hat{\pi})^2}{\sum_i(y_i - \hat{\pi})^2}
$$
### Tjur's coefficient of discrimination
$$ 
R_T^2 = \bar{\pi}_S - \bar{\pi}_F
$$