---
title: "NHANES Data Exploration"
author: "Hallie Rutten"
date: "1/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(haven)

data1 <- read_xpt('https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_DSQIDS.XPT')
#data1 <- read_xpt('NHANES_DietarySupplementUse.XPT')
data1 <- data1 %>% select( SEQN, DSD090, DSD103, DSD122Q, 
  DSDACTSS, DSQICAFF, DSQICALC, DSQICARB, DSQICHL, DSQICHOL, DSQICOPP, 
  DSQIFA, DSQIFDFE, DSQIFIBE, DSQIIRON, DSQIKCAL, DSQILYCO, DSQIMAGN, 
  DSQIMFAT,  DSQINIAC, DSQIPFAT, DSQIPHOS, DSQIPOTA, DSQIPROT, DSQISELE, 
  DSQISFAT, DSQISODI, DSQISUGR, DSQITFAT, DSQIVB1, DSQIVB12, DSQIVB2, DSQIVB6, 
  DSQIVC, DSQIVD, DSQIVK, DSQIZINC)
names(data1)[names(data1) == "DSD090"  ] <- "HowLongTaken"
names(data1)[names(data1) == "DSD103"  ] <- "Past30DaysTaken"
names(data1)[names(data1) == "DSD122Q" ] <- "DosesPerDay"
names(data1)[names(data1) == "DSDACTSS"] <- "ServingSize"
names(data1)[names(data1) == "DSQICAFF"] <- "Caffine" #(mg)
names(data1)[names(data1) == "DSQICALC"] <- "Calcium" #(mg)
names(data1)[names(data1) == "DSQICARB"] <- "Carbohydrate" #(gm)
names(data1)[names(data1) == "DSQICHL" ] <- "Choline" #(mg)
names(data1)[names(data1) == "DSQICHOL"] <- "Cholesterol" #(mg)
names(data1)[names(data1) == "DSQICOPP"] <- "Copper" #(mg)
names(data1)[names(data1) == "DSQIFA"  ] <- "FolicAcid" #(mcg)
names(data1)[names(data1) == "DSQIFDFE"] <- "Folate" #(mcg)
names(data1)[names(data1) == "DSQIFIBE"] <- "DietaryFiber" #(gm)
names(data1)[names(data1) == "DSQIIRON"] <- "Iron" #(mg)
names(data1)[names(data1) == "DSQIKCAL"] <- "Energy" #(kcal)
names(data1)[names(data1) == "DSQILYCO"] <- "Lycopene" #(mcg)
names(data1)[names(data1) == "DSQIMAGN"] <- "Magnesium" #(mg)
names(data1)[names(data1) == "DSQIMFAT"] <- "MonoUnSatFattyAcids" #(gm)
names(data1)[names(data1) == "DSQINIAC"] <- "Niacin" #(mg)
names(data1)[names(data1) == "DSQIPFAT"] <- "PolyUnSatFattyAcids" #(gm)
names(data1)[names(data1) == "DSQIPHOS"] <- "Phosphorus" #(mg)
names(data1)[names(data1) == "DSQIPOTA"] <- "Potassium" #(mg)
names(data1)[names(data1) == "DSQIPROT"] <- "Protein" #(gm)
names(data1)[names(data1) == "DSQISELE"] <- "Selenium" #(mcg)
names(data1)[names(data1) == "DSQISFAT"] <- "SatFattyAcids" #(gm)	
names(data1)[names(data1) == "DSQISODI"] <- "Sodium" #(mg)
names(data1)[names(data1) == "DSQISUGR"] <- "Sugars" #(gm)
names(data1)[names(data1) == "DSQITFAT"] <- "Fat" #(gm)
names(data1)[names(data1) == "DSQIVB1" ] <- "VitaminB1" #(mg)
names(data1)[names(data1) == "DSQIVB12"] <- "VitaminB12" #(mcg)
names(data1)[names(data1) == "DSQIVB2" ] <- "VitaminB2" #(mg)
names(data1)[names(data1) == "DSQIVB6" ] <- "VitaminB6" #(mg)
names(data1)[names(data1) == "DSQIVC"  ] <- "VitaminC" #(mg)
names(data1)[names(data1) == "DSQIVD"  ] <- "VitaminD" #(mcg)
names(data1)[names(data1) == "DSQIVK"  ] <- "VitaminK" #(mcg)
names(data1)[names(data1) == "DSQIZINC"] <- "Zinc" #(mg)
for(i in 6:37){
  # change the measurements from amount per serving to amount per day
  data1[i] = data1[i] * data1$DosesPerDay * data1$ServingSize
}

data2 <- read_xpt('https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_BMX.XPT')
# data2 <- read_xpt('NHANES_BodyMeasures.XPT')
data2 <- data2 %>% select(SEQN, BMXBMI, BMXHT, BMXWAIST, BMXWT)
names(data2)[names(data2) == "BMXBMI"] <- "BMI"
names(data2)[names(data2) == "BMXHT"] <- "Height"
names(data2)[names(data2) == "BMXWAIST"] <- "Waist"
names(data2)[names(data2) == "BMXWT"] <- "Weight"

allData <- merge(data2,data1, by="SEQN")
```

## SOME PHYSICAL GRAPHS (2017-2020 data, pre-pandemic)

Using [NHANES Body measures dataset](https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Examination&Cycle=2017-2020)

```{r, warning = FALSE, message = FALSE}
# just give dat = [whatever your dataset is]
# just make sure they're all numerical, not categorical
dat = data2

# vector of the column indexes (start at 2 bc first column SEQN is irrelevant)
columns = (2:length(dat))

# all combinations of two columns
idx = combn(columns,2)

# make all the graphs
for(i in 1:length(idx[1,])){
  
  # grab the columns and names we want 
  # (could be done inside the plot functions, but it's more readable this way)
  var1 = unlist(dat[ idx[1,i] ])
  name1 = names(dat[ idx[1,i] ])
  var2 = unlist(dat[ idx[2,i] ])
  name2 = names(dat[ idx[2,i] ])
  # the combn function creates a column for each combination
  # but the references are row-major, which is why i comes second
  # enter idx in the console if you want to look at it
  
  # create linear model (up here just so we can stick it on the graph)
  mod <- lm(var2 ~ var1)
  
  # base R scatterplot, changing the labels to read better
  plot(var2~var1, xlab=name1, ylab=name2, main=paste(name2," ~ ",name1) )
  abline(v=mean( var1, na.rm=TRUE ), col="blue")
  abline(h=mean( var2, na.rm=TRUE ), col="blue")
  abline(mod, col="red")
  
  # with tidyverse
  table <- data.frame(name1=var1, name2=var2)
  
  nGroups = 10
  step <- max(table[1], na.rm=TRUE)/nGroups
  
  table <- table %>%
  mutate( group = step*ceiling(name1/step) )
  
  grouped <- table %>%
    group_by( group ) %>%
    summarize( N=n(), Avg=mean(name2, na.rm=TRUE) )
  
  print(
  ggplot(data=dat, aes(x=unlist(dat[idx[1,i]]), y=unlist(dat[idx[2,i]]) )) +
    geom_point() +
    geom_smooth() +
    geom_point(data=grouped, 
               aes(x=group, y=Avg, color="green", size=1), show.legend=FALSE) +
    geom_smooth(method="lm", level=0, color="red") +
    labs(x=name1, y=name2)
  )
  
  print(paste("Summary of", name1, ":"))
  print( summary(var1) )
  cat("\n")
  # give some summary stats
  print(paste("Summary of", name2, ":"))
  print( summary(var2) )
  cat("\n")
  # the group averages
  print(paste("Grouped averages of", name2, ":"))
  print(grouped)
  cat("\n")
  
  # correlation
  cat(name2,"~",name1,"correlation:",cor(var1,var2, use="complete.obs"),"\n\n")
  
  # and the linear model stuff
  print(paste("Linear model:"))
  print( summary(mod) )

}# end graphing loop
```


## Physical graphs by supplements

Body measures dataset vs [dietary supplements dataset](https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Dietary&Cycle=2017-2020):

```{r}

variable <- names(data1[6:37])
vars = length(variable)
BMI <- vector(length = vars)
Height <- vector(length = vars)
Waist <- vector(length = vars)
Weight <- vector(length = vars)

corTable <- data.frame(variable, BMI, Height, Waist, Weight)
for(i in 1:vars){
  c <- which(names(allData)==corTable$variable[i])
  corTable$BMI[i] = cor(allData[2], allData[c], use="complete.obs")
  corTable$Height[i] = cor(allData[3], allData[c], use="complete.obs")
  corTable$Waist[i] = cor(allData[4], allData[c], use="complete.obs")
  corTable$Weight[i] = cor(allData[5], allData[c], use="complete.obs")
}

corTable

corTable[1,]

```

```{r, echo=FALSE}

corBMI <- cor(allData$BMI, allData$Caffine, use="complete.obs")
corHt <- cor(allData$Height, allData$Caffine, use="complete.obs")
corWa <- cor(allData$Waist, allData$Caffine, use="complete.obs")
corWt <- cor(allData$Weight, allData$Caffine, use="complete.obs")

cat("BMI ~ Caffine correlation:", corBMI, "\n\n")
cat("Height ~ Caffine correlation:", corHt, "\n\n")
cat("Waist ~ Caffine correlation:", corWa, "\n\n")
cat("Weight ~ Caffine correlation:", corWt, "\n\n")

```

```{r}

modBMI1 <- lm(allData$BMI ~ allData$Caffine)
modHt1 <- lm(allData$Height ~ allData$Caffine)
modWa1 <- lm(allData$Waist ~ allData$Caffine)
modWt1 <- lm(allData$Weight ~ allData$Caffine)

```

```{r}

plot(allData$BMI ~ allData$Caffine, main="BMI ~ Caffine")
abline(modBMI1, col="red")
plot(allData$Height ~ allData$Caffine, main="Height ~ Caffine")
abline(modHt1, col="red")
plot(allData$Waist ~ allData$Caffine, main="Waist ~ Caffine")
abline(modWa1, col="red")
plot(allData$Weight ~ allData$Caffine, main="Weight ~ Caffine")
abline(modWt1, col="red")

out = which(allData$Caffine > 20000) #3246

( caff <- allData$Caffine[out] )

#cat("\nAn observation of",caff,"mg of caffine is the equivalent of"
#    ,(caff/111),"Red Bulls, which is not a reasonable amount of caffine."
#    ,"This is likely to be a mis-recording of data, so we will exclude it.")

```

An observation of `r caff` mg of caffeine is the equivalent of `r (caff/111)` Red Bulls, which is not a reasonable amount of caffeine. This is likely to be a mis-recording of data, so we will exclude it.

```{r}

modBMI2 <- lm(allData$BMI[-out] ~ allData$Caffine[-out])
modHt2 <- lm(allData$Height[-out] ~ allData$Caffine[-out])
modWa2 <- lm(allData$Waist[-out] ~ allData$Caffine[-out])
modWt2 <- lm(allData$Weight[-out] ~ allData$Caffine[-out])

```

```{r}

plot(allData$BMI[-out] ~ allData$Caffine[-out], main="BMI ~ Caffine")
abline(modBMI1, col="red")
abline(modBMI2, col="blue")
plot(allData$Height[-out] ~ allData$Caffine[-out], main="Height ~ Caffine")
abline(modHt1, col="red")
abline(modHt2, col="blue")
plot(allData$Waist[-out] ~ allData$Caffine[-out], main="Waist ~ Caffine")
abline(modWa1, col="red")
abline(modWa2, col="blue")
plot(allData$Weight[-out] ~ allData$Caffine[-out], main="Weight ~ Caffine")
abline(modWt1, col="red")
abline(modWt2, col="blue")

ix = which(allData$Caffine > 2000)

allData$Caffine[ix]

caff2 <- min(allData$Caffine[ix])

#cat("\n",caff2,"mg of caffine, the smallest of these four extreme observations,"
#    ,"is the equivalent of",(caff2/111),"Red Bulls."
#    ,"This is not a reasonable amount of caffine. These values are likely to be"
#    ,"either mis-recordings of data or outliers, so we will exclude them.")

```

`r caff2` mg of caffine, the smallest of these four extreme observations, is the equivalent of `r (caff2/111)` Red Bulls. This is not a reasonable amount of caffine. These values are likely to be either mis-recordings of data or outliers, so we will exclude them.

```{r}

modBMI3 <- lm(allData$BMI[-ix] ~ allData$Caffine[-ix])
modHt3 <- lm(allData$Height[-ix] ~ allData$Caffine[-ix])
modWa3 <- lm(allData$Waist[-ix] ~ allData$Caffine[-ix])
modWt3 <- lm(allData$Weight[-ix] ~ allData$Caffine[-ix])

```

```{r}

plot(allData$BMI[-ix] ~ allData$Caffine[-ix], main="BMI ~ Caffine")
abline(modBMI1, col="red")
abline(modBMI2, col="blue")
abline(modBMI3, col="green")
plot(allData$Height[-ix] ~ allData$Caffine[-ix], main="Height ~ Caffine")
abline(modHt1, col="red")
abline(modHt2, col="blue")
abline(modHt3, col="green")
plot(allData$Waist[-ix] ~ allData$Caffine[-ix], main="Waist ~ Caffine")
abline(modWa1, col="red")
abline(modWa2, col="blue")
abline(modWa3, col="green")
plot(allData$Weight[-ix] ~ allData$Caffine[-ix], main="Weight ~ Caffine")
abline(modWt1, col="red")
abline(modWt2, col="blue")
abline(modWt3, col="green")

```

```{r, echo=FALSE}

corBMI <- cor(allData$BMI[-ix], allData$Caffine[-ix], use="complete.obs")
corHt <- cor(allData$Height[-ix], allData$Caffine[-ix], use="complete.obs")
corWa <- cor(allData$Waist[-ix], allData$Caffine[-ix], use="complete.obs")
corWt <- cor(allData$Weight[-ix], allData$Caffine[-ix], use="complete.obs")

cat("BMI ~ Caffine adjusted correlation:", corBMI, "\n\n")
cat("Height ~ Caffine adjusted correlation:", corHt, "\n\n")
cat("Waist ~ Caffine adjusted correlation:", corWa, "\n\n")
cat("Weight ~ Caffine adjusted correlation:", corWt, "\n\n")

```

Garbage!!

### which is why looking at the plots to find outliers is important!